Approximation as Differentiation

### 2025-07-02 main/conclusion.tex
Some related work.

\section{Conclusion and Future Work}

\todo{
\begin{itemize}
\item Explainable AI (XAI) techniques like Gradient-weighted Class Activation Mapping
(Grad-CAM)~\cite{selvaraju20} use reverse-mode AD selectively on specific activations, calculating \emph{heat
maps} that highlight input regions contributing to a given classification decision. It might be possible to
compose these techniques with \GPS whilst still preserving useful round-tripping properties; if so this could
be a basis for end-to-end explainability in systems that combine neural networks with traditional symbolic
computation.
\end{itemize}
}

>>>>

\section{Conclusion and Future Work}
\label{sec:conclusion}

\paragraph{Integrating XAI techniques with Galois slicing.} \todo{Sketch what this might look like.}
Explainable AI (XAI) techniques like Gradient-weighted Class Activation Mapping (Grad-CAM)~\cite{selvaraju20}
use reverse-mode AD selectively on specific activations, calculating \emph{heat maps} that highlight input
regions contributing to a given classification decision. It might be possible to compose these techniques with
\GPS whilst still preserving useful round-tripping properties; if so this could be a basis for end-to-end
explainability in systems that combine neural networks with traditional symbolic computation.

### 2025-07-02 main/related-work.tex
Some related work.

Our use of Stable Domain Theory diverges from the traditional aim of modelling infinite or partial data,
however. Instead, we follow a line of work that uses partiality as a qualitative notion of approximation
suitable for provenance and program slicing (discussed in more detail in \secref{related-work:galois-slicing}
below). Paul Taylor’s characterisation of stable functions via local Galois connections on principle downsets
provides the semantic underpinning for the reverse maps used in Galois slicing~\cite{taylor99}. Our work
builds on these ideas by interpreting Galois slicing as a form of differentiable programming, using the
machinery of stable functions and Galois connections to present Galois slicing in a denotational style.

@@

More recent approaches to automatic differentiation have emphasised its semantic foundations.
\citet{elliott18} proposed a categorical model of AD that interprets programs not only as functions enriched
with their derivatives, giving a compositional account of differentiation based on duality and linear maps.
This line of work connects AD to denotational semantics, abstracting over operational concerns and enabling
principled extensions to differentiable programming languages. Vákár and
collaborators~\cite{vákár22,nunes2023} developed the CHAD framework, using Grothendieck constructions over
indexed categories to capture both values and their tangents in a unified semantic structure. These semantic
perspectives shed light on the categorical structure of AD and guide the design of systems that generalise
beyond real analysis, including the application to data provenance and slicing explored in this paper.

@@

\cite{berry79,berry82}

>>>>

Our use of Stable Domain Theory diverges from the traditional aim of modelling infinite or partial data,
however. Instead, we follow a line of work that uses partiality as a qualitative notion of approximation
suitable for provenance and program slicing (discussed in more detail in \secref{related-work:galois-slicing}
below). Paul Taylor’s characterisation of stable functions via local Galois connections on principle downsets
provides the semantic underpinning for the reverse maps used in Galois slicing~\cite{taylor99}. Our work
builds on these ideas by interpreting Galois slicing as a form of differentiable programming, using the
machinery of CHAD to present Galois slicing in a denotational style.

@@

More recent approaches to automatic differentiation have emphasised its semantic foundations.
\citet{elliott18} proposed a categorical model of AD that interprets programs as functions enriched with their
derivatives, giving a compositional account of differentiation based on duality and linear maps. This line of
work connects AD to denotational semantics, abstracting over operational concerns and enabling principled
extensions to differentiable programming languages. Vákár and collaborators~\cite{vákár22,nunes2023} developed
the CHAD framework upon which this paper was based, using Grothendieck constructions over indexed categories
to capture both values and their tangents in a compositional semantic structure. These semantic perspectives
shed light on the categorical structure of AD and guide the design of systems that generalise beyond real
analysis, including the application to data provenance and slicing explored in this paper.

@@

Galois slicing was introduced by \citet{perera12a} as an operational approach to program slicing for pure
functional programs, based on Galois connections between lattices of input and output approximations. A
connection to stable functions in relation to minimal slices for short-circuiting operations was alluded to
in~\citet{perera13}. Subsequent work extended the approach to languages with assignment and
exceptions~\cite{ricciotti17} and concurrent systems, applying Galois slicing to the
$\pi$-calculus~\cite{perera16d}. For the $\pi$-calculus the analysis shifted from functions to transition
relations, considering individual transitions $P \longrightarrow Q$ between configurations $P$ and $Q$ as
analogous to the edge between $x$ and $f(x)$ in the graph of $f$, and building Galois connections between
$\downset{P}$ and $\downset{Q}$. The main difference with the approach presented here is that the earlier work
also computes \emph{program slices}, using approximation lattices that represent partially erased programs; we
discuss this further in \secref{conclusion} below.

More recent work explored Galois slicing for interactive visualisations. \citet{perera22} presented an
approach where slicing operates over Boolean algebras rather than plain lattices. In this setting every Galois
connection $f \dashv g: A \to B$ has a conjugate $g^\circ \dashv f^\circ: B \to A$, where $f^\circ$ denotes
the De Morgan dual $\neg \comp f \comp \neg$~\cite{jonsson51}. The provenance analysis can then be composed
with its own conjugate to obtain a Galois connection which computes \emph{related outputs} (e.g., selecting a
region of a chart and observing the regions of other charts which share data dependencies). \citet{bond25}
revisited this approach using \emph{dynamic dependence graphs} to decouple the derivation of dependency
information from the analyses that make use of it, and observing that to compute the conjugate analysis one
can just use the opposite graph.
